---
title: "exercise-10"
author: "Madina"
format: html
editor: visual
---

Using the {tidyverse} `read_csv()` function, load the “AVONETdataset1.csv” dataset from [this URL](https://github.com/difiore/ada-datasets/blob/main/AVONETdataset1.csv) as a “tibble” named **d**. As discussed in class, this is a recently published dataset that compiles morphological measurements and information on various ecological variables and geographic range data for more than 11,000 species of birds.

```{r}
#| warning: false
library(tidyverse)
library(skimr)
library(dplyr)
f<-"AVONETdataset1.csv"
d<-read_csv(f, col_names = TRUE)

d <- d |>
  select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width,Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass,Habitat, Migration, Trophic.Level, Trophic.Niche,Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, Range.Size)

skim(d)
```

## **Challenge 1**

### **One-Factor ANOVA and Inference**

### **Step 1**

-   Make boxplots of log(**Mass**) in relation to **Trophic.Level** and **Migration** behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. Also, you will want to convert the variable **Migration** (which is scored as a number: “1”, “2”, or “3”) from class numeric to either being classified as a factor or as a character (string) variable.

```{r}
ggplot(data = d |> drop_na(Trophic.Level),aes(x= Trophic.Level, y = log(Mass))) + 
  geom_boxplot() + 
  geom_jitter() 
 
d<- d |> mutate(Migration = as.factor(Migration))

ggplot(data = d |> drop_na(Migration),aes(x= Migration, y = log(Mass))) + 
  geom_boxplot() + 
  geom_jitter() 
```

**Step 2**

-   Run linear models using the `lm()` function to look at the relationship between log(**Mass**) and **Trophic.Level** and between log(**Mass**) and **Migration**.

    ```{r}
    m1<-lm(log(Mass) ~ Trophic.Level, data = d)
    summary(m1)
    m2<-lm(log(Mass) ~ Migration, data = d)
    summary(m2)
    ```

-   Examine the output of the resultant linear models. Is log(**Mass**) associated with either **Trophic.Level** or **Migration**category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

    --- *F statistics is 78.42 is large enough; with p-value being significantly small -\> showing that we reject the null hypotheses; (that all groups have same mass)*

    *so at least one Trophic level group has different mass;*

    ------------------------------------------------------------------------

    ```         
    F-statistic: 144.5 on 2 and 10983 DF,  p-value: < 2.2e-16
    ```

    -   *the F statistic OF regression model of Migration with log(Mass) large enough to reject the null hypothesis!!!*

-   Given the regression coefficients returned for your **Migration** model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories.

    ```{r}
    summary(m2)
    d<- d|> mutate(Migration = relevel(as.factor(Migration), ref = "2"))
    m3<-lm(log(Mass) ~ Migration, data = d)
    summary(m3)
    ```

*The reference level is Migration 1 and with mean 3.77; the Migration2 and Migration3 is slightly have bigger mean value to 0.76 and 0.38.*

*After relevelling with Migration group 2 as a reference level, the other groups showed lower mean value;*

```{r}
plot(data = d, log(Mass) ~ Migration, col = "red")
```

### **Step 3**

-   Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ “significantly” from one another.

```{r}
m3<-lm(log(Mass) ~ Migration, data = d)
pairwise.t.test(log(d$Mass), d$Migration, p.adj = "bonferroni")
m3 <- aov(log(Mass) ~ Migration, data = d)
posthoc <- TukeyHSD(m3, which = "Migration", ordered = TRUE, conf.level = 0.95)
posthoc  
plot(posthoc, xlim = c(0, 1)) 

```

*Migration group 1 and 2 differ significantly!*

### **Step 4**

-   Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(**Mass**) in relation to **Trophic.Level** and calculate a p value for your original F statistic. You can do this either by programming your own permutation test (e.g., by shuffling values for the predictor or response variable among observations and calculating an F statistic for each replicate) or by using the {infer} workflow and setting `calculate(stat="F")`

```{r}
library(infer) 
d<-d|> mutate(logMass = log(Mass))
permuted.F<- d |>
  specify(logMass ~ Trophic.Level) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "F")

original.F <- aov(log(Mass) ~ Trophic.Level, data = d) |> 
  broom::tidy() |> 
  filter(term=="Trophic.Level")
original.F
visualize(permuted.F)
visualize(permuted.F) + shade_p_value(obs_stat = original.F$statistic, direction = "greater")

 
```

## **Challenge 2**

### **Data Wrangling plus One- and Two-Factor ANOVA**

### **Step 1**

-   Create the following two new variables and add them to AVONET dataset:

    -   **Relative beak length**, which you should calculate as the *residual* of log(**Beak.Length_Culmen**) on log(**Mass**).

    -   **Relative tarsus length**, which you should calculate as the *residual* of log(**Tarsus.Length**) on log(**Mass**).

    ```{r}
    beakm<-lm(log(Beak.Length_Culmen) ~ log(Mass), data = d)
    tarsusm<-lm(log(Tarsus.Length) ~ log(Mass), data = d)

    summary(beakm)
    d <- d |> mutate(rel_beak_length = beakm$residuals)
    d <- d |> mutate(rel_tarsus_length = tarsusm$residuals)
    ```

### **Step 2**

-   Make a boxplot or violin plot of your new relative tarsus length variable in relation to **Primary.Lifestyle** and of your new relative beak length variable in relation to **Trophic.Niche**

    ```{r}
    ggplot(data = d ,aes(x= Primary.Lifestyle, y = rel_tarsus_length)) + 
        geom_boxplot() +
      theme(axis.text.x = element_text(angle = 45, hjust=1))

    ggplot(data = d |> drop_na(Trophic.Niche), aes(x = Trophic.Niche, y= rel_beak_length)) + 
        geom_violin() +
        theme(axis.text.x = element_text(angle = 45, hjust=1))
    ggplot(data = d |> drop_na(Trophic.Niche), aes(x = Trophic.Niche, y= rel_beak_length)) + 
        geom_boxplot() +
        theme(axis.text.x = element_text(angle = 45, hjust=1))


    ```

### **Step 3**

-   Run ANOVA analyses to look at the association between geographic range size and the variable **Migration**. You should first drop those observations for which **Migration** is not scored and also look at the distribution of the variable **Range.Size** to decide whether and how it might need to be transformed. Based on the global model, is range size associated with form of migration? How much of the variance in your measure of range size is associated with **Migration** behavior style?

    ```{r}
    migration<- d |> drop_na(Migration)
    hist(migration$Range.Size)
    migration <- migration |> mutate(logRS = log(Range.Size))

    hist(migration$logRS)
    run_an<-lm(data = migration, logRS ~ Migration) 
    summary(run_an) 

    ```

    *Yes the range size is associated with the Migration groups as F statistics and p-values are significant to reject null hypotheses.*

    *While on 8.69% of the variance is explained by this regression model.*

-   Given the regression coefficients returned in the output of the model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories. Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ “significantly” from one another (see [**Module 20**](https://difiore.github.io/ada-2025/20-module.html)).

    *Since on previous steps I relevelled, now th reference level is Migratoin 2 group. Both other migration categories are different than reference level; Migration 1 is lower than Migration2, and Migration 3 is little bit higher than Migration 2.*

    ```{r}

    migration<- migration|> mutate(Migration = relevel(as.factor(Migration), ref = "3"))
    run_an2<-lm(data = migration, logRS ~ Migration) 
    summary(run_an2) 

    m3 <- aov(logRS ~ Migration, data = migration)
    summary(m3)
    mighoc <- TukeyHSD(m3, which = "Migration", ordered = TRUE, conf.level = 0.95)
    mighoc  
    plot(mighoc, xlim = c(0, 5))
    ```

*Both migration categories 1,2 have lower mean logRS than Migration3. There is large significant difference between Categories 3 and 1.*

### **Step 4**

-   Winnow your original data to just consider birds from the Infraorder “Passeriformes” (song birds).

-   Run separate one-factor ANOVA analyses to look at the association between \[1\] relative beak length and **Primary.Lifestyle** and between \[2\] relative beak length and **Trophic.Level**. In doing so…

    -   Make boxplots of response variable by each predictor and by the combination of predictors.

        ```{r}
        passerf <- d |> filter(Order1 == "Passeriformes")

        ggplot(passerf, aes(x = Primary.Lifestyle, y = rel_beak_length, fill = Primary.Lifestyle)) +
          geom_boxplot() +
          labs(title = "Boxplot of Relative Beak Length by Primary Lifestyle")

        ggplot(passerf, aes(x = Trophic.Level, y = rel_beak_length, fill = Trophic.Level)) +
          geom_boxplot() +
          labs(title = "Boxplot of Relative Beak Length by Trophic Level")

        ggplot(passerf, aes(x = interaction(Primary.Lifestyle, Trophic.Level), y = rel_beak_length)) +
          geom_boxplot() +
          labs(title = "Boxplot of Relative Beak Length by Lifestyle and Trophic Level") +
          theme(axis.text.x = element_text(angle = 45, hjust = 1))
        ```

    -   Run linear models for each predictor separately and interpret the model output.

    ```{r}
    pm1<-lm(data = passerf, rel_beak_length ~ Primary.Lifestyle)
    pm2<-lm(data = passerf, rel_beak_length ~ Trophic.Level)
    summary(pm1)
    summary(pm2)
    ```

*The Aerial group's relative beak length is lowest in comparison to other groups in Lifestyle. Incessorial's have highest average relative beak length;*

*Herbivore's relative beak length is smallest in comparison to other categories in Trophic.Level. Omnivore group has highest beak length;*

*F statistics and P-value showing both Predictors's categories have different average beak length; even though only 5\$ person of variation explain the model.*

### **Step 5**

-   Run a two-factor model to look at the association between relative beak length and both **Primary.Lifestyle** and **Trophic.Level** among the passeriforms. Based on the model output, what would you conclude about how relative beak length is related to these two variables?

    ```{r}
    #two_anova<-lm(rel_beak_length ~ Primary.Lifestyle + Trophic.Level, data = passerf)
    #summary(two_anova)
    two_anova<-aov(rel_beak_length ~ Primary.Lifestyle + Trophic.Level, data = passerf)
    summary(two_anova)
    ```

    *The **F-statistics** (138.1 and 201.2) are very high, and **p-values less than 0.01**, meaning both predictors significantly contribute to explaining variance in relative beak length.*

    *So, **18.2% of the total variance in relative beak length is explained by Primary.Lifestyle and 17.7% Trophic.Level**. The remaining **is unexplained**.*

### **Step 6**

-   Finally, run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term. To do this, you should modify your model formula using the colon operator (`:`) to specify the interaction, e.g., relative beak length \~ **Primary.Lifestyle** + **Trophic.Level** + **Primary.Lifestyle:Trophic.Level**. Based on the model output, what would you now conclude about how relative beak length is related to these two variables?

```{r}
#m8<-lm(rel_beak_length ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = passerf)
#summary(m8)
m8<-aov(rel_beak_length ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = passerf)
#m9<-aov(rel_beak_length ~ Primary.Lifestyle*Trophic.Level, data = passerf)
### m8 and m9 same 
summary(m8)

## 
```

*The Sum Sq (8.41) shows how much beak length variation comes from the interaction between Primary.Lifestyle and Trophic.Level. The high F-value (49.26) and very small p-value mean this interaction is significant, so the effect of Primary.Lifestyle on beak length changes depending on Trophic.Level or vice versa.*

### **Step 7**

-   Use the `interaction.plot()` function to visualize the interaction between **Primary.Lifestyle** and **Trophic.Level** (see [**Module 20**](https://difiore.github.io/ada-2025/20-module.html)).

    ```{r}
    interaction.plot(
      x.factor = passerf$Primary.Lifestyle,   
      trace.factor = passerf$Trophic.Level,   
      response = passerf$rel_beak_length,    
      fun = base::mean,                             
      xlab = "Primary Lifestyle",
      ylab = "Mean Relative Beak Length",
      trace.label = "Trophic Level"
    )
    ```

### **Step 8**

In the exercise above, we really did not do any checking with this dataset to see if the data meet the primary assumptions for standard linear regression and ANOVA, which are that variables/residuals within each grouping level are roughly normally distributed and have roughly equal variances. Sample sizes within each grouping level should also be roughly equal. As noted in [**Module 20**](https://difiore.github.io/ada-2025/20-module.html), a general rule of thumb for “equal” variances is to compare the largest and smallest within-grouping level standard deviations and, if this value is less than 2, then it is often reasonable to presume the assumption may not be violated.

Use this approach to see whether variances in across groups in your various models (e.g., for **relative beak length \~ trophic level**) are roughly equal. Additionally, do a visual check of whether observations and model residuals within groups look to be normally distributed.

```{r}
stats <- d |>
    group_by(Trophic.Level) |>
    summarize(`mean(rel_beak_length)` = mean(rel_beak_length), `sd(rel_beak_length)` = sd(rel_beak_length))

max(stats$`sd(rel_beak_length)`)/min(stats$`sd(rel_beak_length)`)
```

*The ratio between maximum and minumim standard deviation is big around 4.8, saying that variances are not homogenous? Then I should not have run standard ANOVA test, or I should do transform data?*

```{r}
stats <- passerf |>
    group_by(Trophic.Level) |>
    summarize(`mean(rel_beak_length)` = mean(rel_beak_length), `sd(rel_beak_length)` = sd(rel_beak_length))

max(stats$`sd(rel_beak_length)`)/min(stats$`sd(rel_beak_length)`)
```

*while is Passeriform data - The ratio between maximum and minumim standard deviation is less than 2, saying that different groups roughly have equal variances.*

```{r}
par(mfrow = c(1, 2))
hist(na.omit(passerf$rel_beak_length[passerf$Trophic.Level == "Carnivore"]), main = "Carnivore", xlab = "Relative beak Length")
qqnorm(na.omit(passerf$rel_beak_length[passerf$Trophic.Level == "Carnivore"]))

#par(mfrow = c(1, 2))
#hist(na.omit(passerf$rel_beak_length[passerf$Trophic.Level == "Herbivore"]), main = "Herbivore", xlab = "Relative beak Length")
#qqnorm(na.omit(passerf$rel_beak_length[passerf$Trophic.Level == "Hervibore"]))

#par(mfrow = c(1, 2))
#hist(na.omit(passerf$rel_beak_length[passerf$Trophic.Level == "Omnivore"]), main = "Omnivore", xlab = "Relative beak Length")
#qqnorm(na.omit(passerf$rel_beak_length[passerf$Trophic.Level == "Omnivore"]))
```
