---
title: "exercise-10"
author: "Madina"
format: html
editor: visual
---

Using the {tidyverse} `read_csv()` function, load the “AVONETdataset1.csv” dataset from [this URL](https://github.com/difiore/ada-datasets/blob/main/AVONETdataset1.csv) as a “tibble” named **d**. As discussed in class, this is a recently published dataset that compiles morphological measurements and information on various ecological variables and geographic range data for more than 11,000 species of birds.

```{r}
#| warning: false
library(tidyverse)
library(skimr)
library(dplyr)
f<-"AVONETdataset1.csv"
d<-read_csv(f, col_names = TRUE)

d <- d |>
  select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width,Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass,Habitat, Migration, Trophic.Level, Trophic.Niche,Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, Range.Size)

skim(d)
```

## **Challenge 1**

### **One-Factor ANOVA and Inference**

### **Step 1**

-   Make boxplots of log(**Mass**) in relation to **Trophic.Level** and **Migration** behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. Also, you will want to convert the variable **Migration** (which is scored as a number: “1”, “2”, or “3”) from class numeric to either being classified as a factor or as a character (string) variable.

```{r}
ggplot(data = d |> drop_na(Trophic.Level),aes(x= Trophic.Level, y = log(Mass))) + 
  geom_boxplot() + 
  geom_jitter() 
 
d<- d |> mutate(Migration = as.factor(Migration))

ggplot(data = d |> drop_na(Migration),aes(x= Migration, y = log(Mass))) + 
  geom_boxplot() + 
  geom_jitter() 
```

**Step 2**

-   Run linear models using the `lm()` function to look at the relationship between log(**Mass**) and **Trophic.Level** and between log(**Mass**) and **Migration**.

    ```{r}
    m1<-lm(log(Mass) ~ Trophic.Level, data = d)
    summary(m1)
    m2<-lm(log(Mass) ~ Migration, data = d)
    summary(m2)
    ```

-   Examine the output of the resultant linear models. Is log(**Mass**) associated with either **Trophic.Level** or **Migration**category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

    --- F statistics is 78.42 is large enough; with p-value being significantly small -\> showing that we reject the null hypotheses; (that all groups have same mass)

-   so at least one Trophic level group has different mass;

    ```{r}
    qf(0.95, df1 = 3, df2 = 11000)
    #p <- 1 - pf(q = 144.5, df1 = 2, df2 = 10983)  # F test
    #p
    #???
    ```

    ------------------------------------------------------------------------

    ```         
    F-statistic: 144.5 on 2 and 10983 DF,  p-value: < 2.2e-16
    ```

    ```{r}
    qf(0.95, df1 = 2, df2 = 10983)
    f <- (summary(m1)$r.squared * (nrow(d) - (ncol(d) - 1) - 1))/((1 - summary(m1)$r.squared) *(ncol(d) - 1))
    f
    1 - pf(q = f, df1 = nrow(d) - (ncol(d) - 1) - 1, df2 = ncol(d) - 1)
    #but there is different f statistics???
    ```

    -   the F statistic OF regression model of Migration with log(Mass) large enough to reject the null hypothesis!!!

-   Given the regression coefficients returned for your **Migration** model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories.

    ```{r}
    summary(m2)
    d<- d|> mutate(Migration = relevel(as.factor(Migration), ref = "2"))
    m3<-lm(log(Mass) ~ Migration, data = d)
    summary(m3)
    ```

The reference level is Migration 1 and with mean 3.77; the Migration2 and Migration3 is slightly have bigger mean value to 0.76 and 0.38.

After releveling with Migration group 2 as a reference level, the other groups showed lower mean value;

```{r}
plot(data = d, log(Mass) ~ Migration, col = "red")
```

### **Step 3**

-   Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ “significantly” from one another.

```{r}
pairwise.t.test(log(d$Mass), d$Migration, p.adj = "bonferroni")
#m <- aov(Migration ~ log(Mass), data = d)
posthoc <- TukeyHSD(m3, which = "Migration", ordered = TRUE, conf.level = 0.95)
posthoc  
plot(posthoc, xlim = c(0, 1))  # xlim set to show zero

```

Migration group 1 and 3 differ significantly!

### **Step 4**

-   Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(**Mass**) in relation to **Trophic.Level** and calculate a p value for your original F statistic. You can do this either by programming your own permutation test (e.g., by shuffling values for the predictor or response variable among observations and calculating an F statistic for each replicate) or by using the {infer} workflow and setting `calculate(stat="F")`

```{r}
library(infer) 
d<-d|> mutate(logMass = log(Mass))
permuted.F<- d |>
  specify(logMass ~ Trophic.Level) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "F")

original.F <- aov(log(Mass) ~ Trophic.Level, data = d) |> 
  broom::tidy() |> 
  filter(term=="Trophic.Level")
original.F
visualize(permuted.F)
visualize(permuted.F) + shade_p_value(obs_stat = original.F$statistic, direction = "greater")

 
```

## **Challenge 2**

### **Data Wrangling plus One- and Two-Factor ANOVA**

### **Step 1**

-   Create the following two new variables and add them to AVONET dataset:

    -   **Relative beak length**, which you should calculate as the *residual* of log(**Beak.Length_Culmen**) on log(**Mass**).

    -   **Relative tarsus length**, which you should calculate as the *residual* of log(**Tarsus.Length**) on log(**Mass**).

    ```{r}
    beakm<-lm(log(Beak.Length_Culmen) ~ log(Mass), data = d)
    tarsusm<-lm(log(Tarsus.Length) ~ log(Mass), data = d)

    summary(beakm)
    d <- d |> mutate(rel_beak_length = beakm$residuals)
    d <- d |> mutate(rel_tarsus_length = tarsusm$residuals)
    ```

### **Step 2**

-   Make a boxplot or violin plot of your new relative tarsus length variable in relation to **Primary.Lifestyle** and of your new relative beak length variable in relation to **Trophic.Niche**

    ```{r}
    ggplot(data = d ,aes(x= Primary.Lifestyle, y = rel_tarsus_length)) + 
        geom_boxplot()

    ggplot(data = d |> drop_na(Trophic.Niche), aes(x = Trophic.Niche, y= rel_beak_length)) + 
        geom_violin()

    ```

### **Step 3**

-   Run ANOVA analyses to look at the association between geographic range size and the variable **Migration**. You should first drop those observations for which **Migration** is not scored and also look at the distribution of the variable **Range.Size** to decide whether and how it might need to be transformed. Based on the global model, is range size associated with form of migration? How much of the variance in your measure of range size is associated with **Migration** behavior style?

-   Given the regression coefficients returned in the output of the model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories. Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ “significantly” from one another (see [**Module 20**](https://difiore.github.io/ada-2025/20-module.html)).
